---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ageutils

<!-- badges: start -->
[![R-CMD-check](https://github.com/TimTaylor/ageutils/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/TimTaylor/ageutils/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

ageutils provides a collection of efficient functions for working with
individual ages and corresponding intervals. These include functions for
efficient conversion from an age to an interval, aggregation of ages with
associated counts in to intervals and the splitting of interval counts based on
specified age distributions.

Note that similar functionality is/was found in the
[ympes](https://cran.r-project.org/package=ympes) package. Going forward, the
intention is that this functionality will be removed from ympes in favour of
this, more focussed, package.

::: {.pkgdown-devel}
The development version of ageutils can be installed from GitHub with:
```{r, eval = FALSE}
remotes::install_github("TimTaylor/ageutils")
```
:::
