---
title: "Utilities for Working with Age Categories"
output:
  markdown::html_format:
vignette: >
  %\VignetteIndexEntry{Utilities for Working with Age Categories}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

ageutils provides a handful of functions for working with age intervals. These
include:

- `breaks_to_interval()`,
- `cut_ages()`,
- `reaggregate_counts`()`, and
- `reaggregate_rates()`.

`breaks_to_interval()` provides a categorisation based on specified breaks which
represent left-hand interval limits. The resultant groupings span from the
minimum break through to a specified `max_upper` and will always be closed on
the left and open on the right. As an example, if `breaks = c(0, 1, 10, 30)` the
interval categories would be [0, 1), [1, 10), [10, 30) and [30, Inf). Ages above
`max_upper` will be returned as NA.

The returned value is as a data frame with 3 entries;
A factor with a character representation of the interval and two columns
representing the numeric values of the corresponding lower (closed) and upper
(open) bounds.

```{r}
library(ageutils)

breaks_to_interval(breaks = c(0L, 1L, 5L, 15L, 25L, 45L, 65L))

breaks_to_interval(breaks = c(1L, 5L, 15L), max_upper = 25L)
```

## `cut_ages()`

`cut_ages()` provides categorisation of ages based on specified breaks which
represent the left-hand interval limits. Categorisation is based on the breaks
and follows the approach of `breaks_to_interval`.

```{r}
cut_ages(ages = 0:9, breaks = c(0L, 1L, 5L, 15L, 25L, 45L, 65L))

cut_ages(1:10, breaks = c(0L, 4L), max_upper = 9L)

x <- cut_ages(1:100, breaks = c(0L, 1L, 5L, 15L, 25L, 45L, 65L))

str(x)

head(x$interval)
```


## `reaggregate_counts()`

`reaggregate`

`split_interval_counts()` splits counts within a age interval in to counts for
individuals years based on a given weighting. Age intervals are specified by
their lower (closed) and upper (open) bounds, i.e. intervals of the form
[lower, upper).

`aggregate_age_counts()` provides aggregation of counts across ages (in years).
It is similar to a `cut()` and `tapply()` pattern but optimised for speed over
flexibility. Groupings are the same as in `cut_ages()` and counts will
be provided across all natural numbers as well as for missing values.


`reaggregate_interval_counts()` is equivalent to, but more efficient than a call
to to `split_interval_counts()` followed by `aggregate_age_counts()`.

The example below shows how it can be used to redistribute counts across a
desired set of age intervals. We use data included in the package that has been
obtained from the 2021 census and modify this based on our desired interval
limits. 

```{r}
# census data
data(pop_dat)
pop_dat

# each row is for the same region so discard for moment
dat <- subset(pop_dat, select = c(age_category, value))

# extract upper and lower bounds
dat <- transform(
    dat,
    lower_bound = as.numeric(sub("\\[([0-9]+), .+)", "\\1", age_category)),
    upper_bound = as.numeric(sub(".+, (.+))", "\\1", age_category))
)

head(dat, n=10)

# recategorise based on ages
with(
    dat,
    reaggregate_interval_counts(
        lower_bounds = lower_bound,
        upper_bounds = upper_bound,
        counts = value,
        breaks = c(0L, 1L, 5L, 15L, 25L, 45L, 65L),
        max_upper = 100L,
        weights = NULL
    )
)
```
